---
title: "Tree Analysis"
author: "Joseph Crockett"
date: "February 5, 2016"
output: html_document
---
```{r global_options, include = F}
library(knitr)
opts_chunk$set(tidy=T, tidy.opts=list(width.cutoff=60))
```
### Homework Assignment 1

#### Objective Statement
Technological advances in LiDAR systems have both increased the accuracy and reduced cost in recent years, however, both system cost and labor required to process data is prohibitive at the biome-scale.  Using LiDAR measurements of Sierra Nevada conifers, our objective is to determine, if any, the relationship between crown spread and tree height.  A positive correlaton would inform a model that utilizes less costly remote sensing systems (i.e. Landsat/ NAIP) to estimate tree height from aerial images.  From such a model, standing carbon can be estimated from conifer biomass (volumes).

#### Methods


### Data

The data are LiDAR measurements of crown height (m) and radii (m) of 176101 conifers in the Sierra Nevada Mountain Range, CA, from a dataset handled by Professor Joshua Viers, UC Merced. Exact location for observations is given as x-y coordinates (m) as per the Universal Coordinate System

### Code

```{r}
#Step 1: Import data
tree_data <- read.csv("~/Desktop/Documents/Writings/Spring 2016/ES 207/Labs/Lab1/Trees.csv", header = T)  #load LiDAR tree measurements into local environment data frame 

dim(tree_data) #Checking formatting of frame.
str(tree_data)
anyDuplicated(tree_data[,2:5]) # Check for duplicated entries
#The formatting seems appropriate, but I will rename column names to avoid conflicts.
colnames(tree_data) <- c("OBJECTID", "xval", "yval", "height", "radii")

save(tree_data, file = "~/Desktop/tree_data.rda")
```


```{r}

#Step 2: Run Data Diognastics
ncol(tree_data)  #running data diagnostics
nrow(tree_data)
head(tree_data)
tail(tree_data)
summary(tree_data) # shows mean, max, min, median, IQR, range, etc. 

library(ggplot2)#plotting histograms using ggplot2, with a function to plot multiple graphs
library(gtable)
library(grid)
library(gridExtra)
library(ggExtra)

p1 <- ggplot(tree_data , aes(x = zval)) + geom_histogram() + theme(axis.text.x =  element_blank()) + ggtitle("Tree Heights (m)")

p2 <- ggplot(tree_data , aes(x = rval)) + geom_histogram() + theme(axis.title.y = element_blank()) + ggtitle("Crown Radii (m)")

#Histograms of log transformed variables
p3 <- ggplot(tree_data , aes(x = zval)) + geom_histogram(bins = 20) + scale_x_log10() + labs(title = "Tree Heights)", x = "log10 z(m)", size = .5)

p4 <- ggplot(tree_data , aes(x = rval)) + geom_histogram(bins = 20) + scale_x_log10()+ theme(axis.title.y = element_blank()) +labs(title = "Crown Radii", x = "log10 r(m)", size = .5)

grid.arrange(p1, p2, p3, p4, ncol= 2)
#Log transformation gives us normal distribution for tree heights and crown radii.
```


Pearson Correlations
```{r}
# Step 3: Examine Correlation Between Variables
# Is the relationship linear? Comparisons of both Pearson and Spearman's rho can reveal linear vs monotonic relationships.
p5 <- ggplot(data = tree_data, aes(x = zval, y = rval)) + geom_point(position = "jitter", size = .5) + geom_smooth(method = glm)+ labs( title = "Tree Height vs. Crown Radii", x = "Tree Height (m)", y = "Crown Radii (m)")

p6 <- ggMarginal(p5, type = "boxplot")

p7 <- ggplot(data = tree_data, aes(x = zval, y = rval)) + geom_point(position = "jitter", size = .5) + geom_smooth(method = glm) + scale_x_log10() + labs( title = "Transformed Tree Height vs. Crown Radii", x = "Tree Height log10(m)", y = "Crown Radii log10(m)")

p8 <- ggMarginal(p7, type = "boxplot")

grid.arrange(p6, p8, ncol = 2, nrow = 1)

#Test for correlation between Tree Height and Crown Radii
#An assumption of Pearson's correlation is that variables must be approximately normally distributed, so log transformation will ensure that this assumption is held.

cor.test(x = log10(tree_data[,4]), y = log10(tree_data[,5]), na.action = na.omit)

cor.test(x = tree_data[,4], y = tree_data[,5], na.action = na.omit)
#Without assuming normality, we can use Spearman's rho test to test correlation between non-parametric variables

cor.test(x = tree_data[,4], y = tree_data[,5], na.action = na.omit, method = "spearman")

#S < P, thus we can assume linearity.
```


```{r}
#Step 4: creating a function

EDA <- function(x){
  #x is a list of values
  min_tmp <- min(x, na.rm = T)
  max_tmp <- max(x, na.rm = T)
  mdn_tmp <- median(x, na.rm = T)
  mn_tmp <- mean(x, na.rm = T)
  rnge_tmp <- range(x, na.rm = T)
  sd_tmp <- sd(x, na.rm = T)
  cv_tmp <- (sd_tmp/mn_tmp) *100
  print(paste("Minimum =", min_tmp, sep = " "))
  print(paste("Maximum =", max_tmp, sep = " "))
  print(paste("Median =", mdn_tmp, sep = " "))
  print(paste("Mean =", mn_tmp, sep = " "))
  print(rnge_tmp)
  print(paste("Standard Deviation =", sd_tmp, sep = " "))
  print(paste("Coefficient of Variance =", cv_tmp, sep = " "))   
  
}
      
```

Lab 1.5

```{r} 
#Step 5: Subsets
tree_data_limit <- subset(tree_data, xval > 708000 )
tree_data_limit <- subset(tree_data_limit, xval < 708200 )
tree_data_limit <- subset(tree_data_limit, yval > 4334000 )
tree_data_limit <- subset(tree_data_limit, yval < 4334200 )
 #subsetting 708000 < x < 708200 and 4334000 < y < 4334200

#By taking this subset, we are examining in greater detail the center of the study area, where we earlier noted a greater concentration of observations.

ncol(tree_data_limit)  #running data diagnostics
nrow(tree_data_limit)
head(tree_data_limit)
tail(tree_data_limit)
summary(tree_data_limit)

p9 <- ggplot(tree_data_limit , aes(x = zval)) + geom_histogram() + theme(axis.text.x =  element_blank()) + ggtitle("Tree Heights (m)")

p10 <- ggplot(tree_data_limit , aes(x = rval)) + geom_histogram() + theme(axis.title.y = element_blank()) + ggtitle("Crown Radii (m)")

#Histograms of log transformed variables
p11 <- ggplot(tree_data_limit , aes(x = zval)) + geom_histogram(bins = 20) + scale_x_log10() + labs(title = "Tree Heights)", x = "log10 z(m)", size = .5)

p12 <- ggplot(tree_data_limit , aes(x = rval)) + geom_histogram(bins = 20) + scale_x_log10()+ theme(axis.title.y = element_blank()) +labs(title = "Crown Radii", x = "log10 r(m)", size = .5)

grid.arrange(p9, p10, p11, p12, ncol= 2)

```

```{r}
#Step 6: Spatial Plotting

#Comparison of observation locations, with color indicating height and dot size indicating crown radii
tree_gg <- ggplot(tree_data, aes(x = xval, y = yval)) + labs(title = "Position of Sierra Nevada Conifers, LiDAR Scan", x = NULL, y = NULL) + theme_bw() + geom_point(aes(size = rval, color = zval), alpha = 9/10, position = "jitter")
                                                                
tree_gg_s <- ggplot(tree_data_limit, aes(x = xval, y = yval)) + labs(title = "Position of Sierra Nevada Conifers, LiDAR Scan, Subset", x = NULL, y = NULL) + theme_bw()  + geom_point(aes(size = rval, color = zval), alpha = 9/10, position = "jitter")

grid.arrange(tree_gg, tree_gg_s, ncol= 2)
```

```{r}
#Step 7: Descriptive Statistics
library(psych)
describe(tree_data_limit)
#The subset has smaller values of mean, median, sd compared to the full dataset.  Standard error has increased.  The minimum conifer height is the same in the subset, but maximum height and minimum and maximum crown radii have contracted.
#The describe function gives the similiar statistics as the summary function (min, max, etc.), and includes skew, kurtosis, standard deviation, standard error, trimmed mean (dropped top and bottom percentage), and absolute deviation from the median.  Describe does not include the quartiles as summary does.
```
```{r}
#Step 8: Regression Modeling


p13 <- ggplot(data = tree_data, aes(x = zval, y = rval)) + geom_point(position = "jitter", size = .5) + geom_smooth(method = lm)+ labs( title = "Tree Height vs. Crown Radii", x = "Tree Height (m)", y = "Crown Radii (m)")

p14 <- ggplot(data = tree_data, aes(x = rval, y = zval)) + geom_point(position = "jitter", size = .5) + geom_smooth(method = lm)+ labs( title = "Crown Radii vs. Tree Height", x = "Crown Radii (m)", y = "Tree Height (m)")


p15 <- ggplot(data = tree_data) + annotate("text",x = 708000, y = 4334000, label = "RVAL", size = 12) + theme_bw() + theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text.y = element_blank(), axis.text.x = element_blank()) 

p16 <- ggplot(data = tree_data) + annotate("text",x = 708000, y = 4334000, label = "ZVAL", size = 12) + theme_bw() + theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text.y = element_blank(), axis.text.x = element_blank()) 

grid.arrange(p13, p15, p16, p14, ncol = 2, nrow = 2)


library(car)

tree_car1 <-lm(zval ~ rval, tree_data_limit)
tree_car2 <-lm(rval ~ zval, tree_data_limit)

#Residuals are the difference between the observed values and the predicted values, i.e between the linear model we predicted and the observation at that x value.  

#Height as a function of crown radii: This seems an unlikely relationship, though useful if our purpose is to model biomass using radii
summary(tree_car1)
residualPlot(tree_car1)
ncvTest(tree_car1)
spreadLevelPlot(tree_car1)
#The residual plot indicates a slightly non random distribution of the residuals.  Without transformation of the variables, a linear model may not be applicable.

#Radii as a function of tree height: a more likely causal relationship.
summary(tree_car2)
residualPlot(tree_car2)
ncvTest(tree_car2)
spreadLevelPlot(tree_car2)

#Homoscedasticity is the assumption that the dependant variable (i.e tree height) influences with similar variance the independent variable across the range of values. As we can see, our subsetted data (comparable to the full dataset), do not exhibit full homoscedascity, and the spread level plot suggests a power transformation to rectify this.

```
```{r}
#Step 9: Outliers

#Detecting outliers in both linear models
outlierTest(tree_car1)
outlierTest(tree_car2)
#These outliers are points that may a significant influence on the model despite their abnormal distance from other observations.  The outliers in the data could be explained by equipment error; strong resource competetion in an area resulting in tall trees with narrow crowns; little resource competition resulting in wide crowns and relatively short trunks; differences in tree species; or conifers that suffered some sort of disturbance that reduced crown radii or height
```

Step 10: Commit to Git

###Results
Plotting conifer positions reveals non-uniform distribution of trees, with high density in the center of the study area and along the southern border, close to what appears to be a path or riverbed. Distributions of tree height and crown radii are skewed right, and so we log transformed the data to better meet the assumptions of normality. Subsequent  Spearman's rho and Pearson's product moment correlation tests indicated that correlation of height vs. radii exists with r-values of .623 and .598 respectively (p-value < 2.2e-16 and < 2.2e-16).  Tests of residuals and homoscedasticity indicate that without transformation, neither a linear model describing crown radii as a function of tree height nor a model describing tree height as a function will adequately model the relationship (suggested power transformations are .54 and .18 respectively)

###Discussion
Our results indicate that modeling eith tree height as a function of crown radii or crown radii as a function of tree height would not completely describe the relationship.  Ideally, we would test a nonlinear model or include more variables to build a multiple regression model.  


###Limitations
The data used is a dense sample of conifers in the Sierras; however, it only describes a km ^2^ area.  If our purpose is to model the relationship between Sierra Nevada conifers height and crown radii, observations from a larger spatial area would capture variations in terrain that may affect that relationship.
