---
title: "Lab2"
author: "Joseph Crockett"
date: "February 11, 2016"
output: html_document
---
Joseph Crockett  
ES 207- Environmental Data Analysis  
February 11, 2016  

Homework Assignment 2  

Objective:
Measurements from four project sites in Northern California are being analyzed to estimate aboveground carbon stock. Based on a subsample of the most frequently occuring tree genera in the data, we will test whether project areas are independant in terms of aboveground carbon stock, proxied by genera frequency. We hypothesize that productivity is based upon individual site characteristics rather than characteristics of specific genera.

Methods:
The data are first corrected for missing or unrealistic values, including unreported genus and dead wood.  A subset of the most frequent genera is tested for independence (Chi-square test) based upon observaton frequency within genera by project site.    
  Data:
  Data from 4 Northern California riparian project sites between 3/20/2012 and 10/14/2013: Heritage Oak Winery, Cosumnes River Preserve, Sacramento River Red Bluff to Hwy 32, and Napa.  Observations and field measurements of woody plant heights, DBH (diameter at breast height).  Samples were identified by genus-Species.
  Code:
  
```{r warning = F, fig.width = 7, fig.height = 4}
  library(ggplot2)
  library(reshape)
  #read data into r
  ripdata <- read.csv("~/Desktop/Documents/Writings/Spring 2016/ES 207/Labs/Lab1/riparian_survey.csv", stringsAsFactors = F)
  
  #summarize data, looking for false values.
  
  summary(ripdata)
  str(ripdata)
 
 #changing Project Codes to simplified codes
 tempo <- unique(ripdata[,2])
 tempn <- c("HOWY","CORP", "SRRB","NASO")
 ripdata$ProjCode <- tempn[match(ripdata[,2], tempo)]
 
 #removing rows with species name = "Not Recorded","Unknown", or "dead wood" or illogical values, like negative values for DBH
 
 unique(ripdata[,10])
 unique(ripdata[,11])
 
 dim(ripdata) # number of rows before data cleansing
 
 ripdata <- ripdata[!(ripdata[,14] == -9999 | ripdata[,15] == -9999 | ripdata[,14] < 0 | ripdata[,10] == "Not Recorded" | ripdata[,10] == "UNKNOWN" | ripdata[,10] == "NODE(UNKNOWN)" | ripdata[,11] == "Unknown" | ripdata[,11] == "dead wood" | ripdata[,11] == "Not Recorded" ), ]
 
which.max(ripdata[,15])
ripdata[3911,]
#A height of 104 m and a dbh of 1.4 cm seems unlikely.  

ripdata <- ripdata[-3911,]
 which.max(ripdata[,15]) 
 ripdata[3270,]
 
 which.max(ripdata[,14]) 
 ripdata[1348,]
 ripdata[617,]
 #229.5 cm DBH and 19 m height seems appropriate.

  which.min(ripdata[,7])
   #The CORP value in the south west seems strange
 ripdata[1349,]
 range(ripdata[ripdata[,16] == "CORP",7])
 
 sum(ripdata[,7] == 36.46)
 #10 points have the same latitude.  This could be a recording error by a researcher or an actual location.  The rest of the points for that site are 38.46. I conclude that changing the values of 36.46 to 38.46 is acceptable.
 
 ripdata$Latitude[ripdata$Latitude == 36.46]  <- 38.46
 dim(ripdata) #number of removed rows after data cleansing
  #We removed 4632 to 3270 rows from the original dataset.  
 
 #Summarize by project code
 ProjLoc <- aggregate(cbind(Longitude, Latitude) ~ ProjCode, data = ripdata, mean)
```

```{r warning = F, fig.width = 7, fig.height = 4}
 #Step 2: Species Selection
 
 #Separating the genera out of the genus - species, and finding the 5 most frequent occurances by project site.
 #SpeciesVarietalName is separted by a space, that can be used to split the terms.
 
 ripdata$Genus <- sapply(strsplit(ripdata$SpeciesVarietalName, " "),"[[", 1)
 
 #freqtable based on project genus ~ site
 genfreq <- table(ripdata[,17],ripdata[,16])
 genfreq
 

#graph by genus ~ site
genfreq_m <- melt(t(genfreq))
ggplot(data = genfreq_m, aes(x = Var.2, y = value, fill = Var.1)) +geom_bar(stat = "identity", position = "stack") + theme_bw() + theme(axis.text.x = element_text(angle = 90)) + labs(title = "Genus by Site", x = "Genus", y = "Count")
 sss <- (aggregate(genfreq_m$value, by = list(Category = genfreq_m$Var.2), FUN = sum))
 head(sss[order(-sss[,2]),], 5)
 #the 5 genus with the greatest frequency are Populus (824), Salix (789), Quercus (703), Fraxinus (467), Acer (457)
 
ripdata_s5 <- subset(ripdata, Genus == "Populus" | Genus == "Salix" | Genus == "Quercus" | Genus == "Fraxinus" | Genus == "Acer",)
```

```{r warning = F, fig.width = 7, fig.height = 4}
 #Step 3: Test for independence
 genfreq_s5 <- table(ripdata_s5[,17],ripdata_s5[,16])
 genfreq_s5
 chisq.test(x = genfreq_s5)
 #A p.value < 2.2e-16 is less than the .05 significance level, we reject the null hypothesis that the subset of genera is independant of site location.
```

```{r warning = F, fig.width = 7, fig.height = 7 } 
 
 #Step 4: Mapping Project location
 ggplot(data = ripdata, aes(x = Longitude, y = Latitude)) + geom_point(aes(color = Genus, shape = ProjectID), alpha = .5) + borders("state", xlim = range(ripdata[,6]), ylim = range(ripdata[,7]))  + theme_minimal() + labs(title = "Site Location and Distribution of Genera", x = "Longitude", y = "Latitude")
   
```
  
```{r warning = F, fig.width = 7, fig.height = 7}
 
 #Step 5: Save df
 write.csv(ripdata_s5, file = "~/Desktop/Documents/Writings/Spring 2016/ES 207/Labs/Lab1/riparian_survey_subset.csv")
 
 #Step 6: Commit to Git
 
``` 
Results:
We removed `r 4632 - 3270` rows from the original dataset due to illogical values or unknown species name, and modified 12 rows to reflect a data input error (latitude were misrecorded as two degrees lower than actual project site latitudes).  By ranking the frequency of observed genera across project sites, we determined that that the five most frequently observed genera are Populus (824), Salix (789), Quercus (703), Fraxinus (467), and Acer (457). A chisquared test of independence on these five genera vs site location leads us to  reject the null hypothesis that frequency is independent of site location (P value < 2.2e-16). 

Discussion:
The result of the chisquared test, that frequency and site location are not independent, reflects what is seen in figure 1.  The Consume River Preserve site had the highest overall proportion of the five most frequent genera, followed by the Sacramento Red River Bluff site.  However, frequency of occurence is not enough for us to estimate differences in carbon stocks.  We could be seeing high occurances of young trees following fire. 

Limitations:
The limitations of a chi squared test are that observations must be independant, expressed in terms of frequency, and with a count ideally greater than 50.  The subset we used met all those criteria: plants could only be one genus, we transformed counts into frequency, and each subsetted genus had more than 50 observations.  However, the chi squared test is unable to tell us anything about the correlation between site area and genus frequency. Further analysis is necessary to determine the relationship, if any.