---
title: "Lab4"
author: "Joseph Crockett"
date: "March 1, 2016"
output: html_document
---
Joseph Crockett
ES 207: Environmental Data Analysis
March 1, 2016
Homework Assignment 4

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, warning = FALSE, message = FALSE)

#check for needed packages, install if needed, require if not
ndpkg <- function(p){
  ifelse(is.element(p,installed.packages()[,1]),require(p,character.only = T), install.packages(p, dep = T))
}
sapply(c("ggplot2", "reshape", "gridExtra", "car", "dplyr", "plyr"), ndpkg)

#load subset data into the global environment as characters
rd_lab4 <- read.csv("~/Desktop/Documents/Writings/Spring 2016/ES 207/Labs/Lab1/riparian_survey_subset.csv", row.names=1, stringsAsFactors=T)
```

Objective Statement:

In order to reasonably estimate standing carbon, we will further refine our linear model between Diameter at Breast Height and Tree height, factored by either genus and Project site.

Methods:

Data:

Code:

```{r Step 1, warnings = F}
rd_lab4$projplot <- as.factor( paste(rd_lab4$ProjCode,rd_lab4$Plot.name))





Results:

Discussion:

Limitations:

