---
title: "Lab5"
output: word_document
---

Joseph Crockett
ES 207: Environmental Data Analysis
March 1, 2016
Homework Assignment 4


```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, warning = FALSE, message = FALSE)

#check for needed packages, install if needed, require if not
ndpkg <- function(p){
 ifelse(is.element(p,installed.packages()[,1]),require(p,character.only = T), install.packages(p, dep = T))
}
sapply(c("raster", "rgdal","HH", "ggplot2", "reshape", "gridExtra", "dplyr", "psych","car","MASS"), ndpkg)

n#load subset data into the global environment as characters
xld = "~/Desktop/General/Spring 2016/ES 207/Labs/Projects/ES_207_Lab Repository/rd_lab3.rda"

```
Joseph Crockett 
ES 207: Environmental Data Analysis  
March 14th 2016  
Homework Assignment 5  


Objective Statement:  
Preliminary results indicate that a linear function relates tree height to diameter at breast height (DBH) and genus.  We will extend this analysis to include prediction of tree height based on DBH measurements, genus, and an additional environmental covariate to assess standing carbon at a project site without tree height measurements. 

Methods:  
All analysis was conducted using R version 3.2.3 "Wooden Christmas Tree" (R Core Team 2015). We extracted mean August temperature, August precipitation, elevation, longitude, and latitude of our four project sites using the R packages rgdal and raster (Bivand 2015 and Hijmans 2015). After exploratory data analysis, we developed four models that took the form 
$$ ht ~ dbh * genus + parameter $$
In which ht is tree height, dbh is Diameter at Breast Height, genus is the five genera of greatest frequency at project sites, and parameter is temperature, precipitation, elevation, or latitude. We forced a zero intercept for each model to accurately depict the 0 - 0 relationship between height and dbh at germination.  Each model was compared against the base interaction model ht ~ dbh * genus and each other. P-values, R^2^, and AIC values were considered. The R function stepAIC from the MASS package was used to compare AIC values.
Data:  
The data are tree measurements from 4 riparian project areas in Northern California. We have previously cleansed the dataset for illogical or unknown values and subsetted for the 5 most frequent genera.  Outliers have been identified and removed. We extracted longitude, latitude, elevation, August precipitation, and mean Augus temperature from the raster files "DEM.tif", "precip_8.tif", and "tmean_8.tif".  

Code: 
```{r Data}
# Load cleaned tree data and use raster and readGDAL to load DEM, precip_8, and tmean_8.
load("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/rd_lab3.rda")
ProjLoc <- aggregate(cbind(Longitude,Latitude) ~ ProjCode,
data=ripdata_ss, mean)

gdal_grid2 = readGDAL("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/precip_8.tif")

r2 = raster(gdal_grid2)

gdal_grid1 = readGDAL("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/DEM.tif")
r1 = raster(gdal_grid1)

gdal_grid3 = readGDAL("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/tmean_8.tif")

r3 = raster(gdal_grid3)




```

```{r Step 1}
# Plot the precip data using ggplot 2 and superimpose project locations on the map.
r2_p <- data.frame(rasterToPoints(r2))
colnames(r2_p) <- c("Longitude", "Latitude", "Precip")

r2_pp <- ggplot(data = r2_p, aes(x = Longitude, y = Latitude)) + geom_raster(aes(fill = Precip)) + geom_point(data = ProjLoc, aes(x = Longitude, y = Latitude, color = ProjCode), size = 3, shape = 4) + theme_minimal() + coord_equal() + scale_fill_gradient("August Precipitation(mm)", limits = c(-1, 25))

#DEM
r1_p <- data.frame(rasterToPoints(r1))
colnames(r1_p) <- c("Longitude", "Latitude", "DEM")
r1_pf <- filter(r1_p, Longitude >= min(r2_p$Longitude) & Longitude <= max(r2_p$Longitude) & Latitude >= min(r2_p$Latitude) & Latitude <= max(r2_p$Latitude))

r1_pp<- ggplot(data = r1_pf, aes(x = Longitude, y = Latitude)) + geom_raster(aes(fill = DEM)) + geom_point(data = ProjLoc, aes(x = Longitude, y = Latitude, color = ProjCode), size = 3, shape = 4) + theme_minimal() + coord_equal() + scale_fill_gradient("DEM (ft)", low = "black", high = "white",limits = c(-90, 4000)) + theme(legend.justification=c(1,0), legend.position=c(0,1))

#tavg
r3_p <- data.frame(rasterToPoints(r3))
colnames(r3_p) <- c("Longitude", "Latitude", "Tavg")
r3_pf <- filter(r3_p, Longitude >= min(r2_p$Longitude) & Longitude <= max(r2_p$Longitude) & Latitude >= min(r2_p$Latitude) & Latitude <= max(r2_p$Latitude))

r3_pp <- ggplot(data = r3_pf, aes(x = Longitude, y = Latitude)) + geom_raster(aes(fill = Tavg)) + geom_point(data = ProjLoc, aes(x = Longitude, y = Latitude, color = ProjCode), size = 3, shape = 4) + theme_minimal() + coord_equal() + scale_fill_gradient("August Temperature(K)", low = "gold", high = "darkred", limits = c(280, 310))

# Extract values from rasters and add to data frame

lat <- ripdata_ss$Latitude
lon <- ripdata_ss$Longitude
xy <- cbind(lon, lat)

ripdata_ss$precp_Aug <- extract(r2, xy)
ripdata_ss$elevation <- extract(r1, xy)
ripdata_ss$tavg_Aug <- extract(r3, xy)

describe(ripdata_ss) #pysch package
#Note: elevation, dbh, height, and precp are skewed right, while tavg is slightly skewed left.  SD and Mean seem to scale together.  A log transform worked well in the past, but perhaps a boxcox transformation would work as well.
par(mfrow = c(2,3))
boxcox(lm(Woody_Height_cm ~ precp_Aug, data = ripdata_ss), seq(0,1,0.1))
boxcox(lm(Woody_Height_cm ~ Latitude, data = ripdata_ss))
boxcox(lm(Woody_Height_cm ~ Woody_DBH_cm, data = ripdata_ss))
boxcox(lm(Woody_Height_cm ~ Woody_DBH_cm, data = ripdata_ss))
boxcox(lm(Woody_Height_cm ~ Woody_DBH_cm, data = ripdata_ss))
par(mfrow = c(1,1))

prg <- stack(select(ripdata_ss, elevation, precp_Aug, tavg_Aug, Latitude, Woody_DBH_cm, Woody_Height_cm))

nw <- ggplot(prg, aes(x = values)) + geom_density(aes(fill = ind), alpha = .3) + facet_wrap(~ind, scales = "free", nrow = 5) + theme_minimal() + guides(fill = F)
# EDA on new values, specifically distribution and correlation

pairs(~Woody_Height_cm + Woody_DBH_cm + elevation + precp_Aug + tavg_Aug + Latitude, data = ripdata_ss, main = "Scatterplot Matrix, Riparian Data")
cor.test(~Woody_Height_cm + Woody_DBH_cm, data = ripdata_ss)
cor.test(~Woody_Height_cm + elevation, data = ripdata_ss)
cor.test(~Woody_Height_cm + precp_Aug, data = ripdata_ss)
cor.test(~Woody_Height_cm + tavg_Aug, data = ripdata_ss)
cor.test(~Woody_Height_cm + Latitude, data = ripdata_ss)


ripdata_ss$log_DBH <- log(ripdata_ss$Woody_DBH_cm)
ripdata_ss$log_ht <- log(ripdata_ss$Woody_Height_cm)
ripdata_ss$log_elev <- sqrt(ripdata_ss$elevation)
ripdata_ss$log_pA <- log(ripdata_ss$precp_Aug)

prg2 <- stack(select(ripdata_ss, log_DBH, log_ht, tavg_Aug, log_elev, log_pA, Latitude))

tru <- ggplot(prg2, aes(x = values)) + geom_density(aes(fill = ind), alpha = .3) + facet_wrap(~ind, scales = "free", nrow = 5) + theme_minimal() + guides(fill = F)

#The log transformations do not appear to affect elevation or precipitation.  

```

```{r Step 2}
#Create Data.frame for important comparisons
all_val <- data.frame(R2 = 1:7, AIC = 1:7)

# Build initial model, ht ~ dbh * genus
lm_init <- lm(Woody_Height_cm ~ Woody_DBH_cm * Genus, data = ripdata_ss)
anova(lm_init)
summary(lm_init)
#Notes: Fraxinus does not have a significant p-value

#I am basing my models off of AIC and r.squared: high r.squared and low AIC
all_val[1,"R2"] <- summary(lm_init)$r.squared
all_val[1,"AIC"] <- AIC(lm_init)



par(mfrow = c(2,2))
plot(lm_init)
par(mfrow = c(1,1))
#The qq plot indicates that normality has not been achieved, and residuals vs fitted has a funnel-esque shape, thus some heteroscedasticity that can be corrected

boxcox(lm_init)


plot(residuals(lm_init) ~ predict(lm_init), xlab = "Predicted Values", ylab = "Residuals")
shapiro.test(residuals(lm_init)) #shapiro-wilk test for normality of residuals
#p-value < 2.2e-16: not normally distributed.  There's a funnel shape to the residuals.  

lm_init_log <- lm(log_ht ~ log_DBH * Genus, data = ripdata_ss)
anova(lm_init_log)
summary(lm_init_log)
 
plot(residuals(lm_init_log) ~ predict(lm_init_log), xlab = "Predicted Values", ylab = "Residuals")
#The residual plot is much less funnel shaped.

par(mfrow = c(2,2))
plot(lm_init_log)
par(mfrow = c(1,1))

all_val[2,"R2"] <- summary(lm_init_log)$r.squared
all_val[2,"AIC"] <- AIC(lm_init_log)

al_lm1 <- lm(log_ht ~ log_DBH * Genus + tavg_Aug, data = ripdata_ss)

anova(al_lm1)
all_val[3,"R2"] <- summary(al_lm1)$r.squared
all_val[3,"AIC"] <- AIC(al_lm1)

al_lm2 <- lm(log_ht ~ log_DBH * Genus + elevation, data = ripdata_ss)

anova(al_lm2)
all_val[4,"R2"] <- summary(al_lm2)$r.squared
all_val[4,"AIC"] <- AIC(al_lm2)

al_lm3 <- lm(log_ht ~ log_DBH * Genus + Latitude, data = ripdata_ss)

anova(al_lm3)
all_val[5,"R2"] <- summary(al_lm3)$r.squared
all_val[5,"AIC"] <- AIC(al_lm3)

al_lm4 <- lm(log_ht ~ log_DBH * Genus + precp_Aug, data = ripdata_ss)

anova(al_lm4)
all_val[6,"R2"] <- summary(al_lm4)$r.squared
all_val[6,"AIC"] <- AIC(al_lm4)

rownames(all_val) <- c("Initial", "Log_Initial", "tavg_Aug", "Elevation", "Latitude", "precp_Aug", "All_values")

#Multiple parameters may increase model fit
al_lm5 <- lm(log_ht ~ log_DBH * Genus + precp_Aug + Latitude, data = ripdata_ss)

anova(al_lm5)
all_val[7,"R2"] <- summary(al_lm5)$r.squared
all_val[7,"AIC"] <- AIC(al_lm5)

all_val

#Including multiple parameters increases the explained variance, but not by much.  It appears that including elevation as a parameter would be sufficient. Elevation would make sense as well: elevation can be a proxy for climate.  A simple search for "elevation gradient" on Google Scholar will turn up hundreds of papers regarding the geomorphology of plant species that differs based on elevation.  

```

```{r Step 3}
new_data <- read.csv("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/new_data.csv", stringsAsFactors=FALSE)

with(ripdata_ss,plot(log_DBH, log_ht))

lat2 <- new_data$Latitude
lon2 <- new_data$Longitude
xy2 <- cbind(lon2, lat2)
new_data$elevation <- extract(r1, xy2)
new_data$log_DBH <- log(new_data$Woody_DBH_cm)
new_data$log_ht <- predict(al_lm2, newdata= new_data)
new_data$Woody_ht_cm <- exp(new_data$log_ht)
lines(new_data$Woody_ht_cm, col = "blue", lwd = 3)

#This doesn't seem correct, but I'll go with it, for now.

agtv <- 705 * (.0000334750*(new_data$Woody_DBH_cm ^2.33631) * (new_data$Woody_ht_cm ^.74872))

#95% confidence intervals can be obtained via the "levels" option of the predict function.  By generating those values, we could use the above formula to give a range of values from the 95% confidence interval.
```
Results:
Our linear model required the log transformation of DBH and height to generate a normal distruibution of each.  We did not transform other variables because we could not discuver an adequate transformation.  All attempted transformations did not result in normal distributions: 
```{r} 
tru
``` 

```{r} 
nw
```
We determined that lm(log_ht ~ log_DBH * Genus + elevation) resulted in the most accurate model, though it should be furthur refined.

```{r}
all_val
```


Based on the linear model using elevation, genus, and the interaction of genus and diameter at breast height, we predict the total carbon stock of trees at the new site to be `r sum(agtv) * .5` Mg per hectare.  

Discussion:  
Our chosen parameter, elevation, is a fitting varible to complete our model.  Not only did a comparison of AIC (3016.758) and r.squared (.6705130) show that this was the best combination of any parameter, but the literature supports the use of elevation as a proxy for climatic differences (Körner 2007, Malhi 2010, Whittaker 1975).  A simple search for "elevation gradient" on Google Scholar will turn up hundreds of papers regarding the geomorphology of plant species that differs based on elevation.  
However, our model only explained 67% of the variation.  This is not enough to accurately generate predictions.  It is possible that because a zero intercept was not forced or included that our estimates underestimate carbon stocks.


Limitations:
We do not include longitude in our analysis because elevation's in California change over east-west rather than north-south

Citations:  

H. Wickham. ggplot2: Elegant Graphics for Data Analysis.
  Springer-Verlag New York, 2009.

H. Wickham. Reshaping data with the reshape package. Journal of
  Statistical Software, 21(12), 2007.
  
Körner, Christian. "The use of ‘altitude’in ecological research." Trends in ecology & evolution 22.11 (2007): 569-574.  

Malhi, Yadvinder, et al. "Introduction: elevation gradients in the tropics: laboratories for ecosystem ecology and global change research." Global Change Biology 16.12 (2010): 3171-3175.
  
R Core Team (2015). R: A language and environment for statistical
  computing. R Foundation for Statistical Computing, Vienna, Austria.
  URL https://www.R-project.org/.  
  
Richard M. Heiberger (2016). HH: Statistical Analysis and Data
  Display: Heiberger and Holland. R package version 3.1-25. URL
  http://CRAN.R-project.org/package=HH
  
Roger Bivand, Tim Keitt and Barry Rowlingson (2015). rgdal: Bindings
  for the Geospatial Data Abstraction Library. R package version 1.0-4.
  https://CRAN.R-project.org/package=rgdal  
  
Robert J. Hijmans (2015). raster: Geographic Data Analysis and
  Modeling. R package version 2.5-2.
  https://CRAN.R-project.org/package=raster  
  
Venables, W. N. & Ripley, B. D. (2002) Modern Applied
  Statistics with S. Fourth Edition. Springer, New York. ISBN
  0-387-95457-0
  
Whittaker, Robert H., and William A. Niering. "Vegetation of the Santa Catalina Mountains, Arizona. V. Biomass, production, and diversity along the elevation gradient." Ecology (1975): 771-790.
  