---
title: "Lab 6"
output: word_document
---

Joseph Crockett
ES 207: Environmental Data Analysis
March 29th, 2016
Homework Assignment 6


```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, warning = FALSE, message = FALSE)

#check for needed packages, install if needed, then adds to environment
ndpkg <- function(p){
 if(!is.element(p,installed.packages()[,1]))
   {install.packages(p, dep = T)}
  require(p,character.only = T)
}
sapply(c("foreign", "ggplot2", "reshape", "raster", "rgdal","dplyr","gridExtra"), ndpkg)

#EDA function
CV1 <- function(x){
  print((sd(x)/mean(x))*100)
}

```


Objective Statement:  

The multitude of services that meadows provide incure special challenges to land managers.  Meadows capture sediment, reduce peak water flows, provide habitat, and have societal and cultural significance.  To guide the decisions of land managers, the US Forest service developed fourteen meadow types based on hydrological and geomorphological characteristics.  We will use

Methods: 
All analysis was conducted in RStudio Version 0.99.891 using R Version 3.2.3 "Wooden Christmas Tree". After removing rows without a corresponding hydogeomorphic type from a the provided Sierra Nevadas meadow polygon layer compiled by Joshua Viers, we further cleaned the dataset by renaming trucated attributes.
  
Data:  
```{r Data}
snmmpc <- read.dbf("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/Sierra_Nevada_MultiSource_Meadow_Polygons_Compilation_v1.dbf")
```
  
Code:  
```{r Step 1: Clean Data}
#Adding ! before a function such as is.na reverses the function; i.e is.na asks whether a value is NA, while !is.na asks whether a value is not NA
summary(snmmpc)
# HGM_TYPE has 16601 NA values

mdwhgm <- snmmpc[!is.na(snmmpc[,"HGM_TYPE"]),] #create new data.frame without NA hgm rows
summary(mdwhgm)

#Important variables to note from the GTR seem to be:  edge.compl, comp_name, Kf, clayTot_r, catchment_area, elev_mean, elev_range,flow_slope, lat, lon, veg_majority

mdwhgm$area_sqkm <- mdwhgm[,"Shape_Area"]/1000000 #m2 to k2
mdwhgm$catch_sqkm <- mdwhgm[,"CATCHMENT_"]/1000000 #m2 to k2
mdwhgm$elev_m <- mdwhgm[,"ELEV_MEAN"] #m
mdwhgm$elev_r <- mdwhgm[,"ELEV_RANGE"] #m
mdwhgm$lat <- mdwhgm[,"LAT_DD"]#decimal degrees
mdwhgm$lon <- mdwhgm[,"LONG_DD"]#decimal degrees
mdwhgm$slope_pct <- mdwhgm[,"FLOW_SLOPE"]
mdwhgm$edge_comp <- mdwhgm[,"EDGE_COMPL"]
mdwhgm$clay <- mdwhgm[,"ClayTot_r"]
mdwhgm$soil_kf <- mdwhgm[,"Kf"]

#EDA
summary(mdwhgm)
sapply(mdwhgm[,29:38],sd) #standard deviation
sapply(mdwhgm[,29:38],function(x) ((sd(x)/mean(x))*100) ) #Coefficient of Variation
#Several variables have maximum values that are orders of magitude greater than the third quarter values: area_sqkm, catch_sqkm, elev_r and to lesser extent, clay, and edge_comp. The coefficient of variance calculations show that though the sd of mean elevation is pretty large, area_sqkm, catch_sqkm, elev_r, and slope_pct have the most variance of the variables. 

#distributions
d <- melt(mdwhgm[,29:38])
d_g <- ggplot(d, aes(x = value)) + facet_wrap(~variable, scales = "free_x") + geom_histogram()
d_g
pairs(mdwhgm[,29:38], col = mdwhgm$HGM_TYPE)

#Histograms of the ten identified variables show mostly right skewed distributions, though mean elevation, latitude, and longitude appear nearly normal.  

#The scatterplots reveal that longitude and latitude have a strong negative correlation, clay and soil_kf have a moderate positive correlation, latitude and mean elevation have a weak negative correlation, and longitude and mean elevation have a weak positive correlation.
rel_cols <-colnames(mdwhgm[,29:38]) #list of variables
```

```{r Step 2: Clustering Output}

#Hierarchical clustering, first finding euclidean distance (staight-line)
md_dist <- dist(x = mdwhgm[,rel_cols], method = "euclidean")
md_hier <- hclust(md_dist, method = "ward.D") #ward.D finds compact, spherical clusters
plot(md_hier)
rect.hclust(md_hier, k = 6)

mdwhgm$hc6 <- as.factor(cutree(md_hier, k = 6))

#K-means clustering
md_km6 <- kmeans(mdwhgm[,rel_cols],centers= 6)
mdwhgm$km6 <- as.factor(md_km6$cluster)

#compare results of each
table(mdwhgm$hc6, mdwhgm$km6)
#It appears that the kmeans clustering and the hierarchial clustering agree on few classifications: groups 4, 5, and 6 saw some agreements, but not many. The following plot displays this mismatch.  Agreements occur primarily between the 40 and 42 parallels.

gdal_grid1 = readGDAL("~/Desktop/General/Spring 2016/ES 207/Labs/Projects/Lab-1/Data/DEM.tif")
r1 = raster(gdal_grid1)

r1_p <- data.frame(rasterToPoints(r1))
colnames(r1_p) <- c("Longitude", "Latitude", "DEM")
r1_pf <- filter(r1_p, Longitude >= min(mdwhgm$lon) & Longitude <= max(mdwhgm$lon) & Latitude >= min(mdwhgm$lat) & Latitude <= max(mdwhgm$lat))

r1_pp<- ggplot(data = r1_pf, aes(x = Longitude, y = Latitude)) + geom_raster(aes(fill = DEM)) + geom_point(data = mdwhgm, aes(x = lon, y = lat, color = hc6), size = 3, shape = 4) + theme_minimal() + coord_equal() + scale_fill_gradient("DEM (ft)", low = "black", high = "white",limits = c(-90, 4000))

r1_pp2<- ggplot(data = r1_pf, aes(x = Longitude, y = Latitude)) + geom_raster(aes(fill = DEM)) + geom_point(data = mdwhgm, aes(x = lon, y = lat, color = km6), size = 3, shape = 4) + theme_minimal() + coord_equal() + scale_fill_gradient("DEM (ft)", low = "black", high = "white",limits = c(-90, 4000))

grid.arrange(r1_pp, r1_pp2, ncol = 2)

```

```{r Step 3: PCA}
#Using Principal Component Analysis to determine the variables explaining the most variance

mdwhgm_pca <- prcomp(x = mdwhgm[,rel_cols], scale = T, retx = T, center = T, scores = T)
plot(mdwhgm_pca, t ="l")
#The first five components explain 85%  of the variance, though an additional 6th explain 90%.  In addition, PC1 and PC2 explain the most variance ( .3469 and .187 respectively) 
 rot <- melt(mdwhgm_pca$rotation)
ggplot(rot, aes(x = as.character(X2), y = value, fill = X1)) + geom_bar(stat = "identity",position = "dodge") + theme_minimal() + labs(title = "PCA Loadings", x = "Principal Component", y = "Variability") 

#In the first five components, strong (greater than .4) positive influence is exerted by latitude, area_sqkm, edge_comp, slope_pct, and catch_sqkm while catch_sqkm, longitude, and soil_kf negatively influences PC4 (less than .4).

biplot(mdwhgm_pca, choices = c(1,2), cex = .5, xlim = c(-.1,.2), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(1,3), cex = .5, xlim = c(-.1,.2), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(1,4), cex = .5, xlim = c(-.1,.2), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(1,5), cex = .5, xlim = c(-.1,.2), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(1,6), cex = .5, xlim = c(-.1,.2), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(2,3), cex = .5, xlim = c(-.1,.3), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(2,4), cex = .5, xlim = c(-.1,.3), ylim = c(-.2,.2))
biplot(mdwhgm_pca, choices = c(2,5), cex = .5, xlim = c(-.1,.3), ylim = c(-.1,.3))
biplot(mdwhgm_pca, choices = c(3,4), cex = .5, xlim = c(-.1,.2), ylim = c(-.2,.2))
biplot(mdwhgm_pca, choices = c(3,5), cex = .5, xlim = c(-.1,.2), ylim = c(-.1,.2))
biplot(mdwhgm_pca, choices = c(4,5), cex = .5, xlim = c(-.15,.2), ylim = c(-.1,.25))

pairs(mdwhgm_pca$x[,1:6],col=mdwhgm$km6)#colored by Kmeans group
pairs(mdwhgm_pca$x[,1:6],col=mdwhgm$hc6)#colored by HClust group

#The functions pairs and biplot will show the same scatterplot of principal component a against principal component b (or multiple pairs in the case of the pairs function), but biplot will also indicate the importance of the loadings.  There are other visual effects, for example the pairs function allows the addition of colors by cluster, while biplot shows the row number of each point, but the loadings of biplot seems to be the key difference.
```
```{r Step 4:Contingency Analysis}
#The kmeans clustering and the hierarchial clustering techniques did not agree on many points, so to perform a contingency analysis, I will compare each technique to the HGM classifications.  
table(mdwhgm$HGM_TYPE, mdwhgm$km6)
table(mdwhgm$HGM_TYPE, mdwhgm$hc6)
#Very few specific relationships appear from these tables.  I would expect a distribution of larger values by row/column if clusters matched classifications. The Annual grassland and Basin peatland classifications have little appearance of relationship to the six kmeans clusters or the hierarchial clusters.  Riparian low gradient appears to have a relationship to at least five of the six hierachial clusters, but no singular correlation.  There seems to be a relationship between subsurface low gradient and kmeans groups 2, 3, and 6, and discharge slope and kmeans 3. Riparian low gradient and kmeans 3 have high counts.
chisq.test(mdwhgm$HGM_TYPE, mdwhgm$hc6)
chisq.test(mdwhgm$HGM_TYPE, mdwhgm$km6)

#Neither clustering technique results in statistically significant (p.value <=  .05) similarity.  

chisq.test(mdwhgm$DOM_ROCKTY, mdwhgm$hc6)
chisq.test(mdwhgm$DOM_ROCKTY, mdwhgm$km6)

chisq.test(mdwhgm$VEG_MAJORI, mdwhgm$hc6)
chisq.test(mdwhgm$VEG_MAJORI, mdwhgm$km6)

table(mdwhgm$DOM_ROCKTY, mdwhgm$hc6)

#However, dominant rock type compares favoribly to both clustering techniques. 
#The lack of overlap between the clustering techniques and the HGM classifications leads me to question whether any conclusions can be made by comparing 6 clusters to 14 classifications.  There appears to be some classifications that are redundant or that have so little relationship to the clusters that they could be dropped (annual grassland, basin peatland, for example). 

```
```{r step 5: ownership}
#Because the clustering results do not match up with any consistency, two tables are needed:

a1 <- aggregate(area_sqkm ~ OWNERSHIP + hc6, data = mdwhgm, FUN = sum, na.action = NULL, na.rm = T)
a2 <- aggregate(area_sqkm ~ OWNERSHIP + km6, data = mdwhgm, FUN = sum, na.action = NULL, na.rm = T)
ggplot(a1,aes(x = OWNERSHIP, y = area_sqkm, fill = as.factor(hc6))) + geom_bar(stat = "identity", position = "dodge") + labs(title = "Total Area, by Ownership and Hierarchial Cluster")+ theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggplot(a2,aes(x = OWNERSHIP, y = area_sqkm, fill = as.factor(km6))) + geom_bar(stat = "identity", position = "dodge") + labs(title = "Total Area, by Ownership and K-mean") + theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 1))

#There does not appear to be a singular cluster that managers could focus on across the National Forests.  However, depending on the clustering, managers could make decisions by forest.  The total area under hierarchial clustering in Inyo National Forest is dominated by cluster 4. Lassen National Forest has a large amount of cluster group 3 by k-means clustering. 

```

Results:  
In depth results are intersperced in the above code, but I will summarize here:  First, ten variables were determined to possibly impact the grouping of riparian zones. Dominant underlying rock formation and majority vegetation, were also considered, but were ultimately not included.  Two clustering functions, hierarchial and k-means, were applied; however, there was little consensus between the two.  Six cluster groups were used for each procedure.  I expected that there would be some overlap between the clusters, but as the tables in step 2 show, there were few direct matchups. Kmean cluster 1 and Hierarchial cluster 5, 3 and 2, 4 and 1, 5 and 3, and 6 and 4 did have high co-tabulation.  
Five principal components further explained ~89% of the variance.  Latitude, area_sqkm, edge_comp, slope_pct, and catch_sqkm positively affected components while  longitude, and soil_kf negatively influenced PC4.  
The six cluster groupings did not compare well to the designated HGM classifications.  Riparian low gradient and riparian middle gradient had   relationships to all groupings, while annual grassland and basin peatland had few cotabulations.  Chi-squared tests further indicated that the hierarchial groupings and classifications were independent (p = .15). K-means clustering and classifications however were statistically similar (p = .0027).  Because we limited our clusters to six groupings rather than the fourteen of the HGM classifications, it is possible that with more clusters, we could more easily refactor the classifications.  The table of kmeans against HGM classification indicates that if we were to remove annual grassland, dry, depressional seasonal, and basin peatland, we could see a better result.  These four classifications have very few cotabulations, thus unnecessary.  
Ideally, our clustering would indicate specific meadow clusters that forest managers could focus on to manage for climate change vulnerability.  However, with the exception of Inyo and Lassen National Forests, cluster groupings are well dispersed within each management area.  

Limitations:  
By attempting to simiplify the classification of meadows, we limited our ability to compare clusters to classifications.  Six groups do not well compare to 14 groups.  In addition, HGM characteristics include vegetation and dominant rock, but I was unable to include either in clustering or PCA.  Finally, the lack of agreement between clustering techniques is some cause for concern, though the k-means clustering did compare favoribly to the HGM classification.

Citations:  
Brinson, Mark M. A hydrogeomorphic classification for wetlands. EAST CAROLINA UNIV GREENVILLE NC, 1993.

