---
title: "Lab7"
output: html_document
---

Joseph Crockett  
ES 207: Environmental Data Analysis  
April 1st, 2016  
Homework Assignment 7: Time & Space  


```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, warning = FALSE, message = FALSE)

#check for needed packages, install if needed, then adds to environment
ndpkg <- function(p){
 if(!is.element(p,installed.packages()[,1]))
   {install.packages(p, dep = T)}
  require(p,character.only = T)
}
sapply(c("foreign", "ggplot2", "reshape", "raster", "rgdal","dplyr","gridExtra"), ndpkg)

#EDA function
CV1 <- function(x){
  print((sd(x)/mean(x))*100)
}

```


Objective Statement:   
Methods:  
Data:  
Code: 
```{r Step 1}
# The provided function getCDEC encounters problems when the date format "m/d/Y" is used: csv files will not be written with the / symbol.  The fixed function, also provided, corrects this with the addition of "as.character(strptime(...))" on lines 53 and 54.  An alternative to this would be to use the function "format" to rearrange the start and end times to "m-d-Y" regardless of the input format.


Results:  
Discussion:  
Limitation:  